---
title: "Homework 2"
author: "Lillian Holden"
date: "2023-09-25"
output: html_document 
---

```{r}
library(manipulate)
```
### Poisson Formula
f(x) =(e– λ λx)/x!

What is the probability that she will hear more than 8 calls during any given session?



What is the probability that she will hear no calls in a session?

P(X=0) = (15^0 e^-15)/0!
= e^-15


What is the probability that she will hear exactly 3 calls in a session?

P(X=3) = (15^3 e^-15)/3!
1125e^-15
3e^-4
5.49% 

## Plot the relevant Poisson mass function over the values in range 0 ≤ x ≤ 30.
```{r}
x <- 0:30
l = 15
probset <- dpois(x = x, lambda = l)
barplot(probset, names.arg = x, space = 0, xlab = "x", ylab = "Pr(X = x)", main = "Probability Mass Function")
```

Simulate 104 results from this distribution (i.e., 2 years of Saturday monitoring sessions).
Plot the simulated results using hist() and use xlim() to set the horizontal limits to be from 0 to 30. How does your histogram compare to the shape of the probability mass function you plotted above?
