---
title: "Modules 6 + 7 Code/Notes"
author: "Lillian Holden"
output: html_document
date: "2023-09-19"
---

##Module 6 

```{r}

install.packages("dplyr")
install.packages("ggplot2")
library(dplyr)
library(ggplot2)
```

##Challenge 1
```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/Country-Data-2016.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
summary(d)
names(d)
```

###Bullet Point One

```{r}
summary(d)
median(d$population, na.rm=TRUE)
median(d$area,na.rm = TRUE)
```


###Bullet Point Two
```{r}
d$density <- d$population/d$area
d <- d[order(-d$density), ]
d[1:10, ]
d <- d[order(d$density), ]
d[1:10, ]
```

###Extracting 20 Largest Countries 
```{r}
d$density <- d$population/d$area
d <- d[order(-d$density), ]
d[1:20, ]
d <- d[order(d$density), ]
d[1:20, ]
median(d[1:20, ]$population, na.rm=TRUE)
median(d[1:20, ]$area, na.rm=TRUE)
```

###Extracting All Countries A-F

```{r}
new <- d[grep("^[A-F]", d$country), ]
summary(new)
mean(new$population, na.rm = TRUE)
mean(new$area, na.rm = TRUE)
```


##Challenge 2
```{r}
par(mfrow = c(2, 3))
boxplot(d$population)
boxplot(log(d$population))
boxplot(d$area)
boxplot(log(d$area))
barplot(d$population)
barplot(d$area)
```

```{r}
par(mfrow = c(1, 2))  # gives us two panels
attach(d)
hist(log(population), freq = FALSE, col = "red", main = "Plot 1", xlab = "log(population size)",
ylab = "density", ylim = c(0, 0.2))
hist(log(area), freq = FALSE, col = "red", main = "Plot 2", xlab = "log(area)",
ylab = "density", ylim = c(0, 0.2))
```

```{r}
v=mean(log(population))
```

```{r}
par(mfrow = c(1, 1))  # set up one panel and redraw the log(population) histogram
hist(log(population), freq = FALSE, col = "white", main = "My Plot with Mean and Density",
xlab = "log(population size)", ylab = "density", ylim = c(0, 0.2))
abline(v = mean(log(population), na.rm = TRUE), col = "blue")
lines(density(log(population), na.rm = TRUE), col = "green")
```

```{r}
detach(d)
```

##Challenge 4
```{r}
sort(table(d$govt_form), decreasing = TRUE)
```

##Challenge 5
```{r}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/KamilarAndCooperData.csv")
d <- read.csv(f, header = TRUE, stringsAsFactors = FALSE)
attach(d)
head(d)
summary(d)
boxplot(log(Body_mass_female_mean) ~ Family, d)
detach(d)
```

```{r}
library(ggplot2)
p <- ggplot(data = d, aes(x = Family, y = log(Body_mass_female_mean)))  #define the variables
p <- p + geom_boxplot()  #graph them in a boxplot
p <- p + theme(axis.text.x = element_text(angle = 90))  #put x-axis names at 90deg
p <- p + ylab("log(Female Body Mass)")  #rename y-axis title
p  #show me the graph
```

--factor levels are meaningful to R, individual data points that have the same variable can be organized based on the same variables--

##Challenge 6
```{r}
attach(d)
par(mfrow = c(1, 2))
plot(x = Body_mass_female_mean, y = Brain_Size_Female_Mean)
plot(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean))
```

```{r}
detach(d)
```

```{r}
p <- ggplot(data = d, aes(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean),
color = factor(Family)))  # first, we build a plot object and color points by Family
p <- p + xlab("log(Female Body Mass)") + ylab("log(Female Brain Size)")  # then we modify the axis labels
p <- p + geom_point()  # then we make a scatterplot
p <- p + theme(legend.position = "bottom", legend.title = element_blank())  # then we modify the legend
p  # and, finally, we plot the object
```

```{r}
p <- p + facet_wrap(~Family, ncol = 4)
p <- p + theme(legend.position = "none")
p
```

```{r}
p <- p + geom_smooth(method = "lm", fullrange = TRUE)
p
```

##Challenge 7
```{r}
p <- ggplot(data = d, aes(x = log(Body_mass_female_mean), y = log(MaxLongevity_m)))
p <- p + geom_point()
p <- p + geom_smooth(method = "lm")
p
```


###Aggregate Statistics and the dplyr package
```{r}
aggregate(d$Body_mass_female_mean ~ d$Family, FUN = "mean", na.rm = TRUE)
```
OR
```{r}
aggregate(x = d["Body_mass_female_mean"], by = d["Family"], FUN = "mean", na.rm = TRUE)
```

```{r}
library(dplyr)
s <- filter(d, Family == "Hominidae" & Mass_Dimorphism > 2)
head(s)  # filtering a data frame for certain rows...
s
```

```{r}
s <- arrange(d, Family, Genus, Body_mass_male_mean)  # rearranging a data frame...
head(s)
s
```

```{r}
s <- select(d, Family, Genus, Body_mass_male_mean)  # selecting specific columns...
head(s)
s
```

```{r}
s <- rename(d, Female_Mass = Body_mass_female_mean)
head(s$Female_Mass)  # renaming columns...
s
```

```{r}
s <- mutate(d, Binomial = paste(Genus, Species, sep = " "))
head(s$Binomial)  # and adding new columns...
s
```

```{r}
s <- summarise(d, avgF = mean(Body_mass_female_mean, na.rm = TRUE), avgM = mean(Body_mass_male_mean,
    na.rm = TRUE))
s
```

```{r}
byFamily <- group_by(d, Family)
byFamily
s <- summarise(byFamily, avgF = mean(Body_mass_female_mean, na.rm = TRUE), avgM = mean(Body_mass_male_mean,
na.rm = TRUE))
s
```

###Piping
```{r}
s <-                                                             #to create dataframe "s"
d %>%                                                          #take dataframe "d"
group_by(Family) %>%                                           #Group it by Family
summarise(avgF = mean(Body_mass_female_mean, na.rm=TRUE),      #And calculate mean male BM
avgM = mean(Body_mass_male_mean, na.rm=TRUE))        #And mean female BM
s
```

##Challenge 8
```{r}
s <- d %>%
mutate(Binomial = paste(Genus, Species, sep = " ")) %>%
select(Binomial, Body_mass_female_mean, Body_mass_male_mean, Mass_Dimorphism) %>%
group_by(Binomial) %>%
summarise(avgF = mean(Body_mass_female_mean, na.rm = TRUE), avgM = mean(Body_mass_male_mean,
na.rm = TRUE), avgBMD = mean(Mass_Dimorphism, na.rm = TRUE))
s
```

###Bullet Point One
```{r}
print(s,n=203)
```

```{r}
s[70:72,]
```

Can do %>% filter Binomial: name of binomial | Binomial name of binomial

Comments:
Vervet monkeys are more sexually dimorphic
Wolly monkeys have a larger average female body mass

###Bullet Point 2

--NEED TO FINISH THIS BULLET POINT

```{r}
s %>% filter(Family == "Cercopithecidae" | Family == "Atelidae) %>% 
```

MAKE SEPERATE PLOTS OF FEMALE AVG, MALE AVG, BM --



##Module 7 


```{r}
install.packages("sciplot")
library(sciplot)
```

##Challenge 1
```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 25, 50, 100, 200, 1000)
gm1 <- function(x) {
prod(x)^(1/length(x))
}
gm1(x)
gm2 <- function(x) {
exp(mean(log(x)))
}
gm2(x)
```

##Challenge 2
```{r}
ss1 <- function(x) {
sum((x - mean(x))^2)
}
ss1(x)
ss2 <- function(x) {
sum(x^2) - length(x) * mean(x)^2
}
ss2(x)
ss3 <- function(x) {
sum(x^2) - (sum(x))^2/length(x)
}
ss3(x)
pop_v <- function(x) {
sum((x - mean(x))^2)/(length(x))
}
pop_v(x)
```

##Challenge 3
```{r}
sample_v <- function(x) {
sum((x - mean(x))^2)/(length(x) - 1)
}
sample_v(x)
var(sample_v(x))
var(sample_v(x), na.rm=TRUE)
sample_v
var(x)
plot(c(0, 50), c(0, 15), type = "n", xlab = "Sample size", ylab = "Variance")
for (n in seq(5, 50, 5)) # samples of 5, 10, 15... for (n in seq(5, 50,
for (n in seq(5, 50, 5)) # samples of 5, 10, 15... 5)) # samples of 5, 10,
for (n in seq(5, 50, 5)) # samples of 5, 10, 15... 15...
{
for (i in 1:50) # 50 replicates for (i in 1:50) # 50 replicates
{
x <- rnorm(n, mean = 10, sd = 2)
points(n, var(x))
}
}
plot(c(0, 50), c(0, 15), type = "n", xlab = "Sample size", ylab = "Variance")
for (n in seq(5, 50, 5)) # samples of 5, 10, 15... for (n in seq(5, 50,
for (n in seq(5, 50, 5)) # samples of 5, 10, 15... 5)) # samples of 5, 10,
for (n in seq(5, 50, 5)) # samples of 5, 10, 15... 15...
{
for (i in 1:50) # 50 replicates for (i in 1:50) # 50 replicates
{
x <- rnorm(n, mean = 10, sd = 2)
points(n, var(x))
}
}
pop_sd <- function(x) {
sqrt(pop_v(x))
}
pop_sd(x)
sample_sd <- function(x) {
sqrt(sample_v(x))
}
sample_sd(x)
```

##Challenge 4

```{r}
SE1 <- function(x) {
sqrt(sample_v(x)/length(x))
}
SE1(x)
var(x)
var(x)/length(x)
sqrt(var(x)/length(x))
library(sciplot)
se(x)
```

###Calculating Confidence Intervals using Standard Errors
```{}
set.seed(1)
x <- rnorm(10000, 0, 1)
hist(x)
x <- seq(from = -4, to = 4, by = 0.01)
plot(x, dnorm(x), cex = 0.4)
plot(x, pnorm(x), cex = 0.4)
x <- seq(from = 0, to = 1, by = 0.01)
plot(qnorm(x), x, cex = 0.4)
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)
m <- mean(x)
n <- length(x)
v <- var(x)
s <- sd(x)
e <- sqrt(v/n)
upper <- mean(x) + qnorm(0.975, mean = 0, sd = 1) * se(x)
lower <- mean(x) + qnorm(0.025, mean = 0, sd = 1) * se(x)  # or lower <- mean(x) - qnorm(0.975)*se(x)
ci <- c(lower, upper)
ci
upper <- m + qnorm(0.975, mean = 0, sd = 1) * e
lower <- m + qnorm(0.025, mean = 0, sd = 1) * e  # or lower <- m - qnorm(0.975)*e
ci <- c(lower, upper)
ci
normalCI = function(x, CIlevel = 0.95) {
upper = m + qnorm(1 - (1 - CIlevel)/2) * sqrt(var(x)/length(x))
lower = m + qnorm((1 - CIlevel)/2) * sqrt(var(x)/length(x))
ci <- c(lower, upper)
return(ci)
}
normalCI(x, 0.95)  # call the function
```

###Calculating Confidence Intervals by Bootstrapping
```{r}
set <- NULL  # sets up a dummy variable to hold our 10000 simulations
n <- 15
for (i in 1:10000) {
set[i] <- mean(sample(x, n, replace = TRUE))
}
quantile(set)
quantile(set, c(0.025, 0.975))
```

##Challenge 5
```{r}
set <- NULL  # sets up a dummy variable to hold our 10000 simulations
n <- 5
for (i in 1:10000) {
set[i] <- mean(sample(x, n, replace = TRUE))
}
quantile(set)
quantile(set, c(0.025, 0.975))
```

```{r}
set <- NULL  # sets up a dummy variable to hold our 10000 simulations
n <- 50
for (i in 1:10000) {
set[i] <- mean(sample(x, n, replace = TRUE))
}
quantile(set)
quantile(set, c(0.025, 0.975))
```

```{r}
set <- NULL  # sets up a dummy variable to hold our 10000 simulations
n <- 500
for (i in 1:10000) {
set[i] <- mean(sample(x, n, replace = TRUE))
}
quantile(set)
quantile(set, c(0.025, 0.975))
```

-- The CI calculated by simulation is reassuringly close to that calculated based on assuming normal distribution. It is not identical, however, due to slightly different values due to the randomization within the simulation. 
--As the number of observations drawn from the sample with replacement (n) decreases, the width of the CI increases. As the number increases, it decreases. 
