---
title: "Module 19"
author: "Lillian Holden"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("statnet")
library(statnet)
```
## Loading Data
```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/bioanth-stats/module-F23-Group1/AnzaEtAl_2021_ggr_weighted.csv")
bonobo_ggr_edgelist <- read.csv(f, header = TRUE, sep = ",")
head(bonobo_ggr_edgelist) # data about bonobos' GG rubbing interactions
```

```{r}
q <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/bioanth-stats/module-F23-Group1/AnzaEtAl_2021_grooming_weighted.csv")
bonobo_groom_edgelist <- read.csv(q, header = TRUE, sep = ",")
head(bonobo_groom_edgelist) # data about bonobos' grooming interactions
```
- What are we looking at in these data sets?
- Who are the actors (also known as senders) vs the receivers?
- In the first 6 rows we are looking at for each data set, who has the most weighted relationship?

- What does this mean?

Answer in class: each row is an edge, each row (edge) has a weight (can be also minutes, whatever you want it to be), nodes are each category, individuals are bonobo

For directed networks, rows indicate the starting node, and columns indicate the receiving node. Undirected networks are symmetric matrices around the diagonal

## Node-Level Variables
```{r}
j <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/bioanth-stats/module-F23-Group1/AnzaEtAl_2021_attributes.csv")
bonobo_attribute <- read.csv(j, header = TRUE, sep = ",")
bonobo_attribute
```
highest ranking is j 

```{r}
bonobo_ggr_net <- as.network.data.frame(bonobo_ggr_edgelist) # turns GG rubbing edge list into a GG rubbing network
bonobo_groom_net<-as.network.data.frame(bonobo_groom_edgelist) # turns grooming edge list into a grooming network

bonobo_ggr_net %v% "rank" <- bonobo_attribute$rank #adds rank variable to GG rubbing network as a vertex attribute
bonobo_ggr_net %v% "age" <- bonobo_attribute$age #adds age variable to GG rubbing network as a vertex attribute
bonobo_ggr_net %v% "group" <- bonobo_attribute$group #adds group variable to GG rubbing network as a vertex attribute
bonobo_groom_net %v% "rank" <- bonobo_attribute$rank #adds rank variable to grooming network as a vertex attribute
bonobo_groom_net %v% "group" <- bonobo_attribute$group #adds group variable to grooming network as a vertex attribute
```
v is for vertices -- another name for nodes
```{r}
summary(bonobo_ggr_net)
```

## Visualizing social network data
````{r}
par(mfrow = c(1, 2))
plot(bonobo_ggr_net, label = "vertex.names", main="GG rubbing") # left plot
box(col = "black")

plot(bonobo_ggr_net, vertex.col = "rank", label = "vertex.names", main="GG rubbing", vertex.cex = 2, edge.label = "GGR.Weight", edge.label.cex = 0.6) # right plot
box(col = "black")
```

```{r}
par(mfrow = c(1, 2))
plot(bonobo_groom_net, label = "vertex.names", main="Grooming") 
box(col = "black")

plot(bonobo_groom_net, vertex.col = "rank", label = "vertex.names", main="Grooming", vertex.cex = 2, edge.label = "weight", edge.label.cex = 0.6)
box(col = "black")
```

## Quadratic Assignment Procedure (QAP)
### 1. Correlation
Do bonobos who groom each other correlate with bonobos who GG rub with one another?

What is our null hypothesis and alternative hypothesis in this question?
```{r}
gcor(bonobo_ggr_net, bonobo_groom_net)
```

```{r}
set.seed(123)
qap_cor <- qaptest(list(bonobo_ggr_net, bonobo_groom_net), # include both network objects in a list
                gcor, # the function you're using is correlation between networks (gcor) 
                g1=1, # use first graph in the list (in this case the GG rubbing network)
                g2=2, # use second graph in the list (in this case the groom network)
                reps = 1000) # number of permutations to run (1000 is actually the default)
summary(qap_cor)
```
```{r}
plot(qap_cor)
```

## Linear/Multiple Regression
```{r}
nodes <- length(bonobo_attribute$actor) # number of nodes in the data set
rank <- bonobo_ggr_net %v% "rank" # a vector of the node-level variable we are interested in

rank_sending <- matrix(data = NA, nrow = nodes, ncol = nodes) # create empty matrix to be filled for senders

for (i in 1:nodes){ # for 1 through the number of nodes in the data set
rank_sending[i,] <- rep(rank[i], nodes)} # the rank of each actor is repeated over entire ROW of matrix
rank_sending
```

```{r}
rank_receiving <- matrix(data = NA, nrow = nodes, ncol = nodes) # empty matrix for receivers

for (i in 1:nodes){ # for 1 through the number of nodes in the data set
rank_receiving[,i] <- rep(rank[i], nodes)} # the rank of each receiver is repeated over entire COLUMN of matrix
rank_receiving
```

```{r}
set.seed(123)
lrqap <- netlm(bonobo_ggr_net, # response variable is a network object with a weighted adjacency matrix
                list(rank_sending)) # list of predictor variables as network or matrix objects
summary(lrqap)
```
```{r}
set.seed(123)
lrqap2 <- netlm(bonobo_ggr_net, # response variable is a network object with a weighted adjacency matrix
                list(rank_receiving)) # list of predictor variables as network or matrix objects
summary(lrqap2)
```

```{r}
set.seed(123)
mrqap <- netlm(bonobo_ggr_net, # response variable is a network object with a weighted adjacency matrix
                list(rank_sending, rank_receiving, bonobo_groom_net)) # list of all predictor variables as network or matrix objects

summary(mrqap)
```

## Logistic regression
```{r}
set.seed(123)
logqap <- netlogit(bonobo_ggr_net, # response variable is a network object
                  list(rank_receiving, rank_sending), # list of all predictor variables as network or matrix objects
                  reps = 1000) # number of draws for quantile estimation, 1000 reps is the default
summary(logqap)
```

## Homophily
```{r}
group <- bonobo_ggr_net %v% "group" # first create a vector giving the group of each node
group_homophily <- outer(group, group, FUN = "==") # shows us whether each combination of terms in vector group is the same (TRUE) or different (FALSE)
group_homophily 
```

```{r}
set.seed(123)
logqap2 <- netlogit(bonobo_ggr_net,
                  list(rank_receiving, rank_sending, group_homophily),
                  reps = 1000)
summary(logqap2)
```

## Spider Monkey Social Networks
```{r}
spmonkey_dat <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/bioanth-stats/module-F23-Group1/spidermonkey_beh_edgelist.csv")
spmonkey_el <- read.csv(spmonkey_dat)
head(spmonkey_el)
```

```{r}
unique(spmonkey_el$Beh) #subsetting the column for proximity types
```

```{r}
mating_spmonkey <- spmonkey_el[spmonkey_el$Beh == "mating",] #making a separate data frame for mating
mating_spmonkey <- mating_spmonkey[,-c(2,4)] #removing unnecessary column data, we'll add back this "Mins" data as edge attributes later 
sp_mating_net <- as.network(mating_spmonkey, matrix.type = "edgelist", directed = TRUE)
plot(sp_mating_net, label = "vertex.names", main="Spider Monkey Mating Network")
```

```{r}
sp_attr <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/bioanth-stats/module-F23-Group1/spidermonkey_attributes.csv")
sp_attr <- read.csv(sp_attr)
head(sp_attr)
```

```{r}
sp_attr$age[sp_attr$age == "infant"] <- 1 #assigning a numeric value to each life history stage
sp_attr$age[sp_attr$age == "juvenile"] <- 2
sp_attr$age[sp_attr$age == "adult"] <- 3
sp_attr$age <- as.numeric(sp_attr$age)
class(sp_attr$age) #confirming the age attribute is coming up as a numeric class
```

```{r}
sp_attr$sex[sp_attr$sex == "female"] <- 0 #assigning a numeric value to each sex
sp_attr$sex[sp_attr$sex == "male"] <- 1
sp_attr$sex <- as.numeric(sp_attr$sex)
class(sp_attr$sex) #confirming the sex attribute is coming up as a numeric class
```

```{r}
#next we need to assign the node attributes to the mating network
sp_mating_net %v% "sex" <- sp_attr$sex #adds sex as node attribute
sp_mating_net %v% "age" <- sp_attr$age #adds age as node attribute
sp_mating_net %v% "repr" <- sp_attr$Repr_status #adds reproductive status as a node attribute
```

```{r}
sp_mating_net %e% "time" <- spmonkey_el[spmonkey_el$Beh == "mating",]$Mins 
summary(sp_mating_net) #can summarize the network like we did for the bonobo GG rubbing network
```

```{r}
play_spmonkey <- spmonkey_el[spmonkey_el$Beh == "play",][,-c(2,4)] #making a separate data frame for play behavior
sp_play_net_unweighted <- as.network(play_spmonkey, matrix.type = "edgelist", directed = TRUE, weighted=FALSE)
plot(sp_play_net_unweighted, label = "vertex.names", main="Spider Monkey Play Network")
```

```{r}
# Adding network and edge attributes to network
sp_play_net_unweighted %v% "sex" <- sp_attr$sex #adds sex as node attribute
sp_play_net_unweighted %v% "age" <- sp_attr$age #adds age as node attribute
sp_play_net_unweighted %v% "repr" <- sp_attr$Repr_status #adds reproductive status as node attribute
```

```{r}
groom_spmonkey <- spmonkey_el[spmonkey_el$Beh == "groom",][,-c(2,4)] #making a separate data frame for grooming behavior
sp_groom_net <- as.network(groom_spmonkey, matrix.type = "edgelist", directed = TRUE, loops = TRUE)
plot(sp_groom_net, label = "vertex.names", main="Spider Monkey Grooming Network")
```

```{r}
# Adding network and edge attributes to the network
sp_groom_net %v% "sex" <- sp_attr$sex #adds sex as node attribute
sp_groom_net %v% "age" <- sp_attr$age #adds age as node attribute
sp_groom_net %v% "repr" <- sp_attr$Repr_status #adds reproductive status as a node attribute

# Let's add the "Mins" from the initial spmonkey edgelist as edge weights
sp_groom_net %e% "time" <- spmonkey_el[spmonkey_el$Beh == "groom",]$Mins 
```

```{r}
contact_spmonkey <- spmonkey_el[spmonkey_el$Beh == "In contact",][,-c(2,4)] #making a separate data frame for contact
sp_contact_net <- as.network(contact_spmonkey, matrix.type = "edgelist", directed = TRUE, loops = TRUE)
plot(sp_contact_net, label = "vertex.names", main="Spider Monkey Contact Network")
```

```{r}
# Adding network and edge attributes to the network
sp_contact_net %v% "sex" <- sp_attr$sex #adds sex as node attribute
sp_contact_net %v% "age" <- sp_attr$age #adds age as node attribute
sp_contact_net %v% "repr" <- sp_attr$Repr_status #adds reproductive status as a node attribute

# Let's add the "Mins" from the initial spmonkey edgelist as edge weights
sp_contact_net %e% "time" <- spmonkey_el[spmonkey_el$Beh == "In contact",]$Mins
```

## Correlation
```{r}
set.seed(123)

# get the correlation value
gcor(sp_groom_net, sp_contact_net)
```

```{r}
# is it significant?
qap_cor <- qaptest(list(sp_groom_net, sp_contact_net), # include both network objects in a list
                gcor, # the function you're using is correlation between networks (gcor) 
                g1=1, # use first graph in the list
                g2=2, # use second graph in the list
                reps = 1000) # number of permutations to run
summary(qap_cor)
```

```{r}
plot(qap_cor)
```

## Linear regression
```{r}
nodes <- length(sp_attr$ID) # the number of nodes in the data set
age <- sp_groom_net %v% "age" # a vector of the node-level variable we are interested in

age_sending <- matrix(data = NA, nrow = nodes, ncol = nodes) #create an empty matrix to be filled

for (i in 1:nodes){ # for 1 through the number of nodes in the data set
age_sending[i,] <- rep(age[i], nodes)} # The age of each actor is repeated over entire ROW of matrix
head(age_sending)
```

```{r}
age_receiving <- matrix(data = NA, nrow = nodes, ncol = nodes)

for (i in 1:nodes){ # for 1 through the number of nodes in the data set
age_receiving[,i] <- rep(age[i], nodes)} # The age of each actor is repeated over entire COLUMN of matrix
head(age_receiving)
```

```{r}
set.seed(123)
lrqap.sp <- netlm(sp_groom_net, # response variable is a network object with a weighted adjacency matrix
                list(age_sending)) # list of all predictor variables as network objects or matrix objects

summary(lrqap.sp)
```

## Logistic Regression
```{r}
nodes <- 16 # the number of nodes in the data set
sex <- sp_play_net_unweighted %v% "sex" # a vector of the node-level variable we are interested in

sex_sending <- matrix(data = NA, nrow = nodes, ncol = nodes) #create an empty matrix to be filled

for (i in 1:nodes){ # for 1 through the number of nodes in the data set
sex_sending[i,] <- rep(sex[i], nodes)} # sex of each actor repeated over entire ROW of matrix
head(sex_sending)
```

```{r}
sex_receiving <- matrix(data = NA, nrow = nodes, ncol = nodes)

for (i in 1:nodes){ # for 1 through the number of nodes in the data set
sex_receiving[,i] <- rep(sex[i], nodes)} # sex of each actor repeated over entire COLUMN of matrix
head(sex_receiving)
```

```{r}
set.seed(123)
lrqap <- netlogit(sp_play_net_unweighted, # response variable is a network object with an unweighted adjacency matrix
                  list(sex_receiving, sex_sending), # list of predictor variables as network or matrix objects
                  reps = 1000) 
summary(lrqap)
```

