---
title: "Module 8"
author: "Lillian Holden"
date: "2023-09-22"
output: html_document 
editor_options: 
  chunk_output_type: console
---

```{r}
library(manipulate)
```


```{r}
outcomes <- c(1, 2, 3, 4, 5, 6)
manipulate(hist(sample(outcomes, n, replace = TRUE), breaks = c(0.5, 1.5, 2.5,
    3.5, 4.5, 5.5, 6.5), probability = TRUE, main = paste("Histogram of Outcomes of ",
    n, " Die Rolls", sep = ""), xlab = "roll", ylab = "probability"), n = slider(0,
    10000, initial = 100, step = 100))
```

## Challenge 1

```{r}
nrolls <- 1000
roll <- function(x) {
    sample(1:6, x, replace = TRUE)
}
two_dice <- roll(nrolls) + roll(nrolls)
hist(two_dice, breaks = c(1.5:12.5), probability = TRUE, main = "Rolling Two Dice",
    xlab = "sum of rolls", ylab = "probability")
```

## Challenge 2

What is the probability that you draw a face card?
-Pr(A) = 12/52
```{r}
12/52
```

What is the probability that you draw a King?
-Pr(A) = 4/52
```{r}
4/52
```

What is the probability that you draw a spade?
-Pr(A) = 13/52
```{r}
13/52
```

What is the probability that you draw a spade given that you draw a face card?
-Pr (A) = face card = 12/52 
-Pr(A|B) = spade given face card = 3/12
-Pr (A ⋂ B) = Pr (A | B) × Pr (B) = 3/12 × 12/52 = 3/52
```{r}
3/52
```

What is the probability that you draw a King given that you draw a face card?
Pr (A) = face card = 12/52 
-Pr(A|B) = King given face card = 4/12
-Pr (A ⋂ B) = Pr (A | B) × Pr (B) = 4/12 × 12/52 = 1/13
```{r}
1/13
```

What is the probability that you draw a card that is both from a red suit (hearts or diamonds) and a face card?
-Pr (A) = red suit = 26/52 = 1/2
Pr (B) = face card = 12/52 =
Pr (A | B) = red suit given face card = 6/12
Pr (A ⋂ B) = Pr (A | B) × Pr (B) = 6/12 × 12/52 = 6/52 = 0.1153846

What is the probability that you draw a card that is either a club or not a face card?
Pr (A) = club = 13/52 = 13/52
Pr (B) = not a face card = 40/52
Pr (A ⋂ B) = club and not a face card = 10/52
Pr (A ⋃ B) = Pr (A) + Pr (B) - Pr (A ⋂ B) = 13/52 + 40/52 - 10/52 = 43/52

## Flipping A Fair Coin

```{r}
outcomes <- c("heads", "tails")
prob <- c(1/2, 1/2)
barplot(prob, ylim = c(0, 0.6), names.arg = outcomes, space = 0.1, xlab = "outcome",
    ylab = "Pr(X = outcome)", main = "Probability Mass Function")
```

```{r}
cumprob <- cumsum(prob)
barplot(cumprob, names.arg = outcomes, space = 0.1, xlab = "outcome", ylab = "Cumulative Pr(X)",
    main = "Cumulative Probability")
```
## Rolling a Fair Die
```{r}
outcomes <- c(1, 2, 3, 4, 5, 6)
prob <- c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)
barplot(prob, ylim = c(0, 0.5), names.arg = outcomes, space = 0.1, xlab = "outcome",
    ylab = "Pr(X = outcome)", main = "Probability Mass Function")
```    
## An Example

```{r}
cumprob <- cumsum(prob)
barplot(cumprob, names.arg = outcomes, space = 0.1, xlab = "outcome", ylab = "Cumulative Pr(X)",
    main = "Cumulative Probability")
```

```{r}
library(ggplot2)
a <- 2
b <- 1
K <- 2
x <- seq(from = 0, to = 1, by = 0.025)
fx <- K * x^(a - 1) * (1 - x)^(b - 1)
lower_x <- seq(from = -0.25, to = 0, by = 0.025)  # add some values of x less than zero
upper_x <- seq(from = 1, to = 1.25, by = 0.025)  # add some values of x greater than one
lower_fx <- rep(0, 11)  # add fx=0 values to x<0
upper_fx <- rep(0, 11)  # add fx=0 values to x>1
x <- c(lower_x, x, upper_x)  # paste xs together
fx <- c(lower_fx, fx, upper_fx)  # paste fxs together
d <- as.data.frame(cbind(x, fx))
p <- ggplot(data = d, aes(x = x, y = fx)) + xlab("x") + ylab("f(x)") + geom_line()
p
```

```{r}
library(manipulate)
manipulate(ggplot(data = d, aes(x = x, y = fx)) + xlab("x") + ylab("f(x)") +
    geom_line() + geom_polygon(data = data.frame(xvals = c(0, n, n, 0), fxvals = c(0,
    K * n^(a - 1) * (1 - n)^(b - 1), 0, 0)), aes(x = xvals, y = fxvals)) + ggtitle(paste("Area Under Function = ",
    0.5 * n * K * n^(a - 1) * (1 - n)^(b - 1), sep = " ")), n = slider(0, 1,
    initial = 0.5, step = 0.01))
```
```{r}
pbeta(0.75, 2, 1)  # cumulative probability for x ≤ 0.75
```
```{r}
x <- seq(from = 0, to = 1, by = 0.005)
prob <- 0.5 * x * K * x^(a - 1) * (1 - x)^(b - 1)
barplot(prob, names.arg = x, space = 0, main = "Cumulative Probability", xlab = "x",
    ylab = "Pr(X ≤ x)")
```

```{r}
pbeta(0.5, 2, 1)  # cumulative probability for x ≤ 0.50
```

```{r}
pbeta(0.7, 2, 1)  # yields .49
```

```{r}
qbeta(0.49, 2, 1)  # yield 0.7
```

## Expected Mean and Variance of Random Variables

```{r}
m <- sum(seq(1:6) * 1/6)
m
```

```{r}
var <- sum((seq(1:6) - mean(seq(1:6)))^2 * (1/6))
var
```
### Challenge 3

```{r}
Prob <- (13/52)^1 * (39/52)^0
Prob
```

```{r}
Var <- (13/52)*(1-(13/52))
Var
```


## Challenge 4

```{r}
n <- 6  # number of trials
k <- 6  # number of successes
p <- 1/6
prob <- (factorial(n)/(factorial(k) * factorial(n - k))) * (p^k) * (1 - p)^(n -
    k)
prob
```

```{r}
k <- 3  # number of successes
prob <- (factorial(n)/(factorial(k) * factorial(n - k))) * (p^k) * (1 - p)^(n -
    k)
prob
```

```{r}
dbinom(x = k, size = n, prob = p)
```

```{r}
probset <- dbinom(x = 0:6, size = 6, prob = 1/6)  # x is number of successes, size is number of trials
barplot(probset, names.arg = 0:6, space = 0, xlab = "outcome", ylab = "Pr(X = outcome)",
    main = "Probability Mass Function")
```

```{r}
cumprob = cumsum(probset)
barplot(cumprob, names.arg = 0:6, space = 0.1, xlab = "outcome", ylab = "Cumulative Pr(X)",
    main = "Cumulative Probability")
```

```{r}
sum(probset)  # equals 1, as it should
```

```{r}
dbinom(x = 3, size = 6, prob = 1/6)
```

```{r}
pbinom(q = 3, size = 6, prob = 1/6)  # note the name of the argument is q not x
```

```{r}
sum(dbinom(x = 0:3, size = 6, prob = 1/6))  # this sums the probabilities of 0, 1, 2, and 3 successes
```

```{r}
1 - pnbinom(q = 3, size = 6, prob = 1/6)
```

```{r}
pnbinom(q = 3, size = 6, prob = 1/6, lower.tail = FALSE)
```

```{r}
1 - pbinom(q = 2, size = 6, prob = 1/6)  # note here that the q argument is '2'
```

```{r}
pbinom(q = 2, size = 6, prob = 1/6, lower.tail = FALSE)
```

```{r}
x <- 0:10
l = 3.5
probset <- dpois(x = x, lambda = l)
barplot(probset, names.arg = x, space = 0, xlab = "x", ylab = "Pr(X = x)", main = "Probability Mass Function")
```

```{r}
x <- 0:20
l = 10
probset <- dpois(x = x, lambda = l)
barplot(probset, names.arg = x, space = 0, xlab = "x", ylab = "Pr(X = x)", main = "Probability Mass Function")
```

```{r}
x <- 0:50
l = 20
probset <- dpois(x = x, lambda = l)
barplot(probset, names.arg = x, space = 0, xlab = "x", ylab = "Pr(X = x)", main = "Probability Mass Function")
```

```{r}
x <- 0:10
l <- 3.5
barplot(ppois(q = x, lambda = l), ylim = 0:1, space = 0, names.arg = x, xlab = "x",
    ylab = "Pr(X ≤ x)", main = "Cumulative Probability")
```

```{r}
x <- 0:20
l <- 10
barplot(ppois(q = x, lambda = l), ylim = 0:1, space = 0, names.arg = x, xlab = "x",
    ylab = "Pr(X ≤ x)", main = "Cumulative Probability")
```

```{r}
x <- 0:50
l <- 20
barplot(ppois(q = x, lambda = l), ylim = 0:1, space = 0, names.arg = x, xlab = "x",
    ylab = "Pr(X ≤ x)", main = "Cumulative Probability")
```

DO HOMEWORK PROBLEM HERE

## Challenge 5

```{r}
a <- 4
b <- 8
x <- seq(from = a - (b - a), to = b + (b - a), by = 0.01)
fx <- dunif(x, min = a, max = b)  # dunif() evaluates the density at each x
plot(x, fx, type = "l", xlab = "x", ylab = "f(x)", main = "Probability Density Function")
```

```{r}
plot(x, punif(q = x, min = a, max = b), type = "l", xlab = "x", ylab = "Pr(X ≤ x)",
    main = "Cumulative Probability")  # punif() is the cumulative probability density up to a given x
```

## Challenge 6

```{r}
mu <- 4
sigma <- 1.5
curve(dnorm(x, mu, sigma), mu - 4 * sigma, mu + 4 * sigma, main = "Normal Curve",
    xlab = "x", ylab = "f(x)")
```

```{r}
manipulate(plot(seq(from = (mu - 4 * sigma), to = (mu + 4 * sigma), length.out = 1000),
    dnorm(seq(from = (mu - 4 * sigma), to = (mu + 4 * sigma), length.out = 1000),
        mean = mu, sd = sigma), type = "l", xlim = c(mu - 4 * sigma, mu + 4 *
        sigma), xlab = "x", ylab = "f(x)", main = "Normal Probability Density Function") +
    polygon(rbind(c(mu - nsigma * sigma, 0), cbind(seq(from = (mu - nsigma *
        sigma), to = (mu + nsigma * sigma), length.out = 1000), dnorm(seq(from = (mu -
        nsigma * sigma), to = (mu + nsigma * sigma), length.out = 1000), mean = mu,
        sd = sigma)), c(mu + nsigma * sigma, 0)), border = NA, col = "salmon") +
    abline(v = mu, col = "blue") + abline(h = 0) + abline(v = c(mu - nsigma *
    sigma, mu + nsigma * sigma), col = "salmon"), mu = slider(-10, 10, initial = 0,
    step = 0.25), sigma = slider(0.25, 4, initial = 1, step = 0.25), nsigma = slider(0,
    4, initial = 0, step = 0.25))
```

```{r}
manipulate(plot(seq(from = (mu - 6 * sigma), to = (mu + 6 * sigma), length.out = 1000),
    pnorm(seq(from = (mu - 6 * sigma), to = (mu + 6 * sigma), length.out = 1000),
        mean = mu, sd = sigma), type = "l", xlim = c(-20, 20), xlab = "x", ylab = "f(x)",
    main = "Cumulative Probability"), mu = slider(-10, 10, initial = 0, step = 0.25),
    sigma = slider(0.25, 10, initial = 1, step = 0.25))  # plots the cumulative distribution function
```

```{r}
p <- pnorm(8, mean = 6, sd = 2) - pnorm(7, mean = 6, sd = 2)
p
```

```{r}
mu <- 0
sigma <- 1
p <- pnorm(mu + 2 * sigma, mean = mu, sd = sigma) - pnorm(mu - 2 * sigma, mean = mu,
    sd = sigma)
p
```

```{r}
p <- pnorm(mu + 1 * sigma, mean = mu, sd = sigma) - pnorm(mu - 1 * sigma, mean = mu,
    sd = sigma)
p
```

```{r}
manipulate(plot(seq(from = (mu - 4 * sigma), to = (mu + 4 * sigma), length.out = 1000),
    dnorm(seq(from = (mu - 4 * sigma), to = (mu + 4 * sigma), length.out = 1000),
        mean = mu, sd = sigma), type = "l", xlim = c(mu - 4 * sigma, mu + 4 *
        sigma), xlab = "x", ylab = "f(x)", main = "Normal Probability Density Function") +
    abline(v = mu, col = "blue") + abline(h = 0) + polygon(x = c(qnorm((1 -
    CI)/2, mean = mu, sd = sigma), qnorm((1 - CI)/2, mean = mu, sd = sigma),
    qnorm(1 - (1 - CI)/2, mean = mu, sd = sigma), qnorm(1 - (1 - CI)/2, mean = mu,
        sd = sigma)), y = c(0, 1, 1, 0), border = "red"), mu = slider(-10, 10,
    initial = 0, step = 0.25), sigma = slider(0.25, 10, initial = 1, step = 0.25),
    CI = slider(0.5, 0.99, initial = 0.9, step = 0.01))
```

## Challenge 7

```{r}
n <- 1000
mu <- 3.5
sigma <- 4
v <- rnorm(n, mu, sigma)
mean(v)
```

```{r}
var(v)
```

```{r}
sd(v)
```

```{r}
hist(v, breaks = seq(from = -15, to = 20, by = 0.5), probability = TRUE)
```

```{r}
qqnorm(v, main = "Normal QQ plot random normal variables")
qqline(v, col = "gray")
```

```{r}
p <- ppoints(length(v))
head(p)
```

```{r}
tail(p)
```

```{r}
theoretical_q <- qnorm(ppoints(length(v)))
```

```{r}
observed_q <- quantile(v, ppoints(v))
```

```{r}
plot(theoretical_q, observed_q, main = "Normal QQ plot random normal variables",
    xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
```

## Challenge 8 

```{r}
x <- rnorm(10000, mean = 5, sd = 8)  # simulate from a normal distribution with mean 5 and SD 8
hist(x)
```

```{r}
mean(x)
```

```{r}
sd(x)
```

```{r}
z <- (x - mean(x))/sd(x)  # standardized!
hist(z)
```

```{r}
mean(z)
```

```{r}
sd(z)
```

```{r}
set.seed(1)
x <- rnorm(1e+06, 25, 5)
hist(x, probability = TRUE)
```

```{r}
mu <- mean(x)
mu
```

```{r}
sigma <- sqrt(sum((x - mean(x))^2)/length(x))
```

```{r}
k <- 1000  # number of samples
n <- 5  # size of each sample
s <- NULL  # dummy variable to hold each sample
for (i in 1:k) {
    s[[i]] <- sample(x, size = n, replace = FALSE)
}
head(s)
```

```{r}
m <- NULL
for (i in 1:k) {
    m[i] <- mean(s[[i]])
}
mean(m)  # almost equal to...
```

```{r}
mu
```

```{r}
pop_se <- sqrt(sigma^2/n)
pop_se  # SE estimated from population standard deviation
```

```{r}
pop_se <- sigma/sqrt(n)
pop_se  # SE estimated from population standard deviation
```

```{r}
stdev <- NULL
for (i in 1:k) {
    stdev[i] <- sd(s[[i]])
}
sem <- stdev/sqrt(n)  # a vector of SEs estimated from each sample 
head(sem)
```

```{r}
mean(sem)  # which is almost equal to...
```

```{r}
pop_se
```
